Exploring the tree of numerical semigroups

### 2014-09-02 Papier/article.tex
article.tex

Though there are asymptotically faster algorithm than the one presented here,
thank to careful optimizations, we were able to compute $n_g$ for much larger
genus than before. This is due to the fact that our algorithm is very well
suited for the current processor architecture. In particular, it allows to use
parallelism at various scale (parallel branch exploration, vectorization)...

To get the greatest speed from modern processor we used several optimization
tricks, which we will elaborate in the following section:
\begin{itemize}
\item Vectorization (MMX, SSE instructions sets) and careful memory alignment.
\item Shared memory multi-core computing using \texttt{CILK++} for low level
  enumerating tree branching;
\item Un-recursived algorithm using a stack;
\item Avoiding all dynamic allocation during the computation: everything is
  computed ``in place'';
\item Avoiding all unnecessary copy (the cost of the \textsc{Son} algorithm is
  roughly the same (even less) as for as copy !
\item Aggressive loop unrolling. The main loop is unrolled by hand using some
  kind of Duff's device.
\item Carefull choice of data type (\verb|uint_fast8_t| for decomposition
  number, vs \verb|uint_fast64_t| for all indexes).
\end{itemize}

@@

Assume for example that we want to construct the tree $\mathcal{T}^{\leq 100}$ of all  numerical semigroup of  genus smaller than $100$. 
In this case, the representation of  numerical  semigroup given in Section~\ref{S:DecNumber} uses decomposition numbers of integers smaller than~$300$.
By Lemma~\ref{L:RepN}, such a decomposition number is smaller than $151$ and requires~$1$ byte of memory.
Thus at each \textbf{for} step of Algorithm~\textsc{Son}, the CPU actually works on $1$ byte. 
However current CPUs work on $8$ bytes.
The first optimization uses this point.

@@

\noindent In this version we can see that the cursors \texttt{src} and
\texttt{dst} move at the same time and that the modification of the value
pointed by \texttt{dst} only needs to access the values pointed by \tt{src}
and \tt{dst}.  We can therefore work in multiple entries at the same time
without collision.  Current CPUs allow this thanks to the SIMD technologies as
MMX, SSE, etc.  The acronym SIMD stands for Single Operation Multiple Data.
We used SSE4.1 technology as it allows for the largest speedup. This need to
respect some constraints in the memory organization of the data, namely
``memory alignment''. Recall that an address is 16~bytes aligned if it is a
multiple of 16. SSE memory access are much faster for aligned access.

@@

Then the core of the \textbf{while} loop in the preceding algorithm is
translated as a for loop as follows (recall that $x$ denotes the generator if the
father which is to be removed in the children):
\begin{listing}{1}
start = x >> 4;      // index of the block containing x
shift = x & 0xF;     // offset of x inside the block
...                  // some specific instructions to
...                  // handle the first incomplete block.
for (long int i=start+1; i<NBLOCKS; i++) {
  block = load_unaligned_epi8(src + ((i-start)<<4) - shift);
  dst[i] -= ((block != zero) & one);
}
\end{listing}
The instruction \tt{load\_unaligned\_epi8} (specific to SSE technology) loads
in the variable \tt{block} 16 consecutive entries of the decomposition number
of the parent (called \tt{src} monoid). Those entries will be used to compute
the entries $16i, \dots 16i+15$ of the children monoid. Since the removed
generator $x$ is not necessarily a multiple of $16$, the data are not aligned
in memory, hence the use of a specific instruction. The \tt{zero}
(resp. \tt{one}) constants are initialized as 16 bytes equal to $0$
(resp. $1$). The comparison \tt{(block != zero)} therefore returns a block
which contains $0$ in the bytes corresponding the $0$ entries of block and
$255$ in the non zero one. This results is then bitwise anded with one so that
the instruction actually performs a 16~bytes parallel version of
\begin{algorithmic}
\State $\tt{dst} \gets \tt{dst} -
\textbf{ if } \tt{block} \neq 0 \textbf{ then } 1 \textbf{ else } 0$
\end{algorithmic}
As we previously said, to gain more speed this core loop is actually unrolled
using some kind of Duff device as
\begin{algorithmic}
\State \textbf{Jump to case } $\tt{start}+1$
\State \textbf{case 1:} \tt{block = ...; dst[1] -= ((block != zero) \& one)}
\State \textbf{case 2:} \tt{block = ...; dst[2] -= ((block != zero) \& one)}
\State \textbf{[...]}
\State \textbf{case NBLOCK-1:} \tt{...}
\end{algorithmic}

@@

Our second optimization is to use parallelism on exploration of the tree.
Today, CPU of personal computer have several cores (2, 4 or more).  The given
version of our exploration algorithm uses a single core and so a fraction only
of the power of a CPU. The idea here is that different branches of the tree can
be explored in parallel by different cores of the computer. The tricky part is
to ensure that each cores are busy, giving a new branch when a core is done
with a former one. Fortunately there is a technology calle \texttt{Cilk+}
which is particularly well suited for those kind of problems.

@@

The thing are actually only a little bit more complicated. First we have to
gather the results of the exploration. If we simply write
\begin{algorithmic}[1]
\State \tt{result}[$g(\tt{S}$] $\gets \tt{result}[g(\tt{S}] + 1$
\end{algorithmic}
then we face the problem of to core incrementing the same variable at the same
time. This is called a data race and leads to nondeterministic wrong
result. Cilk provide the notion of \emph{reducer} which are variables local to
each thread which are gathered (here added) when a thread is finishing its
job.

A more important problem is that the cost of a recursive call in non
negligible. Using \tt{Cilk+} recursive calls makes it even worse. The solution
we use is to switch back to the non recursive version using a stack when the
genus is close to the target genus. This leads to the following \tt{Cilk+} code:
\begin{listing}{1}
void explore(const monoid &m) {
  unsigned long int nbr = 0;
  if (m.genus < MAX_GENUS - STACK_BOUND) {
    auto it = generator_iter<CHILDREN>(m);
    while (it.move_next()) {
      auto child = remove_generator(m, it.get_gen()).
      cilk_spawn explore(child);
      nbr++;
    }
    cilk_results[m.genus] += nbr;
  }
  else explore_stack(m, cilk_results.get_array());
}
\end{listing}
Note that in our version, we found that \tt{STACK\_BOUND} optimal value was
around 10-12 for genus un the range $45\dots 67$ so that \tt{explore\_stack}
is used more than $99\%$ of the time, \tt{Cilk+} is used to balance the work
between the different cores.

@@

Avoiding all dynamic allocation during the computation: everything is computed
``in place'';

% Our method of parallelism is very simple: for $G\in \NN$,  we cut the tree $\mathcal{T}^{\leq G}$ in sub-trees $\mathcal{T}_1,...,\mathcal{T}_n$ and we launch our algorithm on these sub-trees. 
% The advantage of this method is that there is no communication between the instances of our algorithm. 
% The disadvantage is that the cutting controls the efficiency of the parallelism. 
% Assume for example that we want to explore the tree~$\mathcal{T}^{\leq 40}$.
% We first determine all numerical semigroups $S_1,...S_{n_{20}}$ of genus $20$. 
% To explore $\mathcal{T}$ it remains to explore the tree $\mathcal{T}_i$ rooted in $S_i$ for $i=1,...,n_{20}$. 
% This works but the time to explore $\mathcal{T}_1$ is similar to the time need to explore the tree $\mathcal{T}^{\leq 40}$. 
% And in this case, using many cores does not reduce the time to explore the tree.

% Let us now explain in more detail how we cut the tree $\mathcal{T}^{\leq G}$ in order to use parallelism.
% Semigroups of the form $\left<g+1,...,2g+1\right>$, which are of genus $g$, are called \emph{ordinary} in~\cite{Elizalde}. 
% Each ordinary semigroup has a unique son that is also ordinary. 
% We define $X$ to be the set of all the non-ordinary sons of an ordinary semigroup and $X^{\leq G}$ the restriction of $X$ to semigroup of genus $\leq G$. 
% We then denote by  $\mathcal{T}_i$ the tree rooted on $S_i$ where $X^{\leq G}=\{S_1,...,S_n\}$. 
% The time needed to explore $\mathcal{T}_i$ is very heterogeneous but there are many tree $T_i$ with maximal time. 
% This cutting is more efficient than the previous one.

>>>>

Though there are asymptotically faster algorithms than the one presented here,
thank to careful optimizations, we were able to compute $n_g$ for much larger
genus than before. This is due to the fact that our algorithm is particularly
well suited for the current processor architecture. In particular, it allows
to use parallelism at various scale (parallel branch exploration,
vectorization)...

To get the greatest speed from modern processor, we used several optimization
tricks, which we will elaborate in the following section:
\begin{itemize}
\item Vectorization (MMX, SSE instructions sets) and careful memory alignment.
\item Shared memory multi-core computing using \texttt{CILK++} for low level
  enumerating tree branching;
\item Partially derecursived algorithm using a stack;
\item Avoiding all dynamic allocation during the computation: everything is
  computed ``in place'';
\item Avoiding all unnecessary copy (the cost of the \textsc{Son} algorithm is
  roughly the same as copying) !
\item Aggressive loop unrolling: the main loop is unrolled by hand using some
  kind of Duff's device.
\item Carefull choice of data type (\verb|uint_fast8_t| for decomposition
  number, vs \verb|uint_fast64_t| for all indexes).
\end{itemize}

@@

Assume for example that we want to construct the tree $\mathcal{T}^{\leq 100}$
of all numerical semigroup of genus smaller than $100$.  In this case, the
representation of numerical semigroup given in Section~\ref{S:DecNumber} uses
decomposition numbers of integers smaller than~$300$.  By Lemma~\ref{L:RepN},
such a decomposition number is smaller than $151$ and requires~$1$ byte of
memory.  Thus at each \textbf{for} step of Algorithm~\textsc{Son}, the CPU
actually works on $1$ byte. However current CPUs usually work on $8$ bytes and
even on $16$~bytes using vector extensions. The first optimization uses this
point.

@@

\noindent In this version we can see that the cursors \texttt{src} and
\texttt{dst} move at the same time and that the modification of the value
pointed by \texttt{dst} only needs to access the values pointed by \tt{src}
and \tt{dst}.  We can therefore work in multiple entries at the same time
without collision.  Current CPUs allow this thanks to the SIMD technologies as
MMX, SSE, etc.  The acronym SIMD~\cite{SIMD} stands for Single Operation
Multiple Data.  We used SSE4.1~\cite{SSE, IntelSSE} technology as it allows
for the largest speedup. This need to respect some constraints in the memory
organization of the data, namely ``memory alignment''. Recall that an address
is 16~bytes aligned if it is a multiple of 16. SSE memory access are much
faster for aligned access.

@@

Then the core of the \textbf{while} loop in the preceding algorithm is
translated as a for loop as follows (recall that $x$ denotes the generator of
the father which is to be removed in the children):
\begin{listing}{1}
start = x >> 4;      // index of the block containing x
shift = x & 0xF;     // offset of x inside the block
...                  // some specific instructions to
...                  // handle the first incomplete block.
for (long int i=start+1; i<NBLOCKS; i++) {
  block = load_unaligned_epi8(src + ((i-start)<<4) - shift);
  dst[i] -= ((block != zero) & one);
}
\end{listing}
The instruction \tt{load\_unaligned\_epi8} (specific to SSE technology) loads
in the variable \tt{block} 16 consecutive entries of the decomposition number
of the parent (called \tt{src} monoid). Those entries will be used to compute
the entries $16i, \dots 16i+15$ of the children monoid. Since the removed
generator $x$ is not necessarily a multiple of $16$, the data are not aligned
in memory, hence the use of a specific instruction. The \tt{zero}
(resp. \tt{one}) constants are initialized as 16 bytes equal to $0$
(resp. $1$). The comparison \tt{(block != zero)} therefore returns a block
which contains $0$ in the bytes corresponding the $0$ entries of block and
$255$ in the non zero one. This results is then bitwise anded with one so that
the instruction actually performs a 16~bytes parallel version of
\begin{algorithmic}
\State $\tt{dst} \gets \tt{dst} -
\textbf{ if } \tt{block} \neq 0 \textbf{ then } 1 \textbf{ else } 0$
\end{algorithmic}
which is equivalent to Lines \tt{10} to \tt{14} of Algorithm~\textsc{Son}.
\medskip

As we previously said, to gain more speed this core loop is actually unrolled
using some kind of Duff device~\cite{Duff} as
\begin{algorithmic}
\State \textbf{Jump to case } $\tt{start}+1$
\State \textbf{case 1:} \tt{block = ...; dst[1] -= ((block != zero) \& one)}
\State \textbf{case 2:} \tt{block = ...; dst[2] -= ((block != zero) \& one)}
\State \textbf{[...]}
\State \textbf{case NBLOCK-1:} \tt{...}
\end{algorithmic}

@@

Our second optimization is to use parallelism on exploration of the tree.
Today, CPU of personal computer have several cores (2, 4 or more).  The given
version of our exploration algorithm uses a single core and so a fraction only
of the power of a CPU. The idea here is that different branches of the tree can
be explored in parallel by different cores of the computer. The tricky part is
to ensure that each cores are busy, giving a new branch when a core is done
with a former one. Fortunately there is a technology called
\texttt{Cilk+}~\cite{CilkIntel, CilkRefman}
which is particularly well suited for those kind of problems.

@@

The thing are actually only a little bit more complicated. First we have to
gather the results of the exploration. If we simply write
\begin{algorithmic}[1]
\State \tt{result}[$g(\tt{S}$] $\gets \tt{result}[g(\tt{S}] + 1$
\end{algorithmic}
then we face the problem of two core incrementing the same variable at the
same time. Incrementing a variables is actually done in 3 steps: reading the
value from the memory, adding one, storing back the results. Since there is by
default non synchronization, the following sequence of actions for two core
and is possible: Read1 / Read2 / Add1 / Add2 / Store1 / Store2. Then the two
cores performs the same modification resulting in incrementing the variable
only once. This is called a data race and leads to nondeterministic wrong
result. To cope with those kinds of synchronization problems, Cilk provide the
notion of \emph{reducer} which are variables local to each thread which are
gathered (here added) when a thread is finishing its job.

A more important problem is that the cost of a recursive call in non
negligible. Using \tt{Cilk+} recursive calls instead of \tt{C++} calls makes
it even worse. The solution we use is to switch back to the non recursive
version using a stack when the genus is close to the target genus. This leads
to the following \tt{Cilk+} code:
\begin{listing}{1}
void explore(const monoid &m) {
  unsigned long int nbr = 0;
  if (m.genus < MAX_GENUS - STACK_BOUND) {
    auto it = generator_iter<CHILDREN>(m);
    while (it.move_next()) {
      auto child = remove_generator(m, it.get_gen()).
      cilk_spawn explore(child);
      nbr++;
    }
    cilk_results[m.genus] += nbr;
  }
  else explore_stack(m, cilk_results.get_array());
}
\end{listing}
Note that in our version, we found that \tt{STACK\_BOUND} optimal value was
around 10-12 for genus in the range $45\dots 67$ so that \tt{explore\_stack}
is used more than $99\%$ of the time. The \tt{Cilk+} recursive function does
actually very little work but ensure that the work is balanced between the
different cores.

@@

Using vectorization and loop enrolling as described previously leads to an
extremely fast \textsc{Son} algorithm. Indeed, its cost is comparable to the cost
of copying a monoid. It is therefore crucial for performance to avoid any
extra cost. We list here various places where unnecessary cost can be avoided.

\subsubsection*{Avoiding dynamic allocation} Compared to the \textsc{Son}
algorithm, dynamic allocation cost orders of magnitude more. Therefore, during
the derecursived stack algorithm, we only allocate (on the system stack rather
than on the heap), the stack of monoid. No further allocation are done.

\subsubsection*{Avoiding all unnecessary copy} We also used a trick to avoid
copying from and to the top of the stack. Indeed, then main loop perform the
following sequences of operations:
\begin{algorithmic}[1]
  \State $\tt{S} \gets \tt{stack.top()}$
  \State \tt{stack.pop()}
  \For{all children $\tt{S}^\tt{x}$ of \tt{S}}
  \State $\tt{S.push}(\tt{S}^\tt{x})$
  \EndFor
\end{algorithmic}
If we use a stack of monoid, we can construct $\tt{S}^\tt{x}$ directly into
the stack memory but we have to copy the top of the stack to \tt{S}. Since the
average number of children is the golden ratio, this single copy in far from
being negligible. The trick is to use a level of indirection, replacing the
stack by an array of monoid and an array of indexes pointing to the array of
monoid. We keep the invariant that the array of indexes is a permutation of
the array of monoid. Now instead of copying \tt{S} out of the stack, we keep
it on the stack, pushing the children in the second position by exchanging the
indexes. Here is the relevant part of the code:
\begin{listing}{1}
  monoid data[MAX_GENUS-1], *stack[MAX_GENUS], *current;
  monoid **stack_pointer = stack + 1;
  for (ind_t i=1; i<MAX_GENUS; i++) stack[i] = &(data[i-1]);
  [...]
  while (stack_pointer != stack) {
      --stack_pointer;
      current = *stack_pointer;
      [...] for each children {
        *stack_pointer = *(stack_pointer + 1);
        [...] construct the children in **stack_pointer
        [...] using the parent in *current
        stack_pointer++;
        [...]
      }
      *stack_pointer = current;
  }
\end{listing}

\subsubsection*{Pointer arithmetic and indexes}

Due to the way, the C++ does its pointer arithmetic, even if the index in the
array are less than $3G$ and therefore fits in $8$~bits, we use $64$~bits
indexes (namely \verb|uint_fast64_t|) to avoid conversion and sign extension
when computing addresses of indexed element. This single standard trick save
20\% of speed.

\subsection{Timings}
